### 2018-03-13 paper/1-intro.tex

Build systems (such as \Make) are big, complicated, but unloved part of
the software ecosystem.  Every developer on the planet uses one, but
they are very much a means to an end, and seldom the focus of
attention.  But the challenges of scale have driven large software firms
like Microsoft, Facebook, and Google to develop their own build
systems, each with its own choices and idiosyncrasies.

@@

In this paper we offer a general framework in which to discuss these
questions in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code).  Specifically we make
these contributions:
\begin{itemize}
\item We describe some simple but novel abstractions that
crisply encapsulate what a build system is.
\item We show that we can instantiate
these abstractions to describe the essence of a variety of different
build systems, including \Make, \Shake, \Bazel, and \Excel, each in
a dozen lines of code or so.
Doing this in a single setting allows
the differences and similarities between these huge systems to be
brought out clearly.
\item Build systems vary on many axes;
for example: static vs dynamic dependencies; cloud-build, including
shallow vs deep; deterministic vs non-deterministic build rules;
early cut-off; self-tracking build systems; and persistent metadata.
These properties (which we define in~\S\ref{sec-background}) are often
deeply-built-in assumptions of a particular build system.
In contrast, our framework allows them to be distinguished,
reasoned about, and varied.
\item We show that all build system we look at can be broken along two axes - the order in which they build dependencies, and whether they decide to build a particular dependency. Interestingly, these two aspects can be recombined without restriction to produce new build systems.
\item Two particularly desirable properties are dynamic dependencies
and cloud-build; yet no currently-available build system supports
both.  Our framework makes it rather easy to do so, for the first
time.
\end{itemize}
Thus equipped, instead of seeing build systems as unrelated
points in space, we can re-envisage them as locations in a landscape,
leading to better understanding of what they do and how they compare,
and suggesting exploration of other (as yet unoccupied points) in the
landscape.

>>>>

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.  Despite this, the challenges of scale have driven
large software firms like Microsoft, Facebook, and Google to develop
their own build systems, each with its own choices and idiosyncrasies.

@@

In this paper we offer a general framework in which to discuss these
questions in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code).  Specifically we make
these contributions:
\begin{itemize}
\item We describe some simple but novel abstractions that
crisply encapsulate what a build system is.
\item We show that we can instantiate
these abstractions to describe the essence of a variety of different
build systems, including \Make, \Shake, \Bazel, and \Excel, each in
a dozen lines of code or so.
Doing this in a single setting allows
the differences and similarities between these huge systems to be
brought out clearly.
\item Build systems vary on many axes,
including: static vs dynamic dependencies; cloud-build, including
shallow vs deep; deterministic vs non-deterministic build rules;
early cut-off; self-tracking build systems; and persistent metadata.
These properties (which we define in~\S\ref{sec-background}) are often
deeply built into any particular build system.
In contrast, our framework allows them to be distinguished,
reasoned about, and varied.
\item We show that all build system we look at can be broken along two axes - the order in which they build dependencies, and whether they decide to build a particular dependency. Interestingly, these two aspects can be recombined without restriction to produce new build systems.
\item Two particularly desirable properties are dynamic dependencies
and cloud-build; yet no currently-available build system supports
both.  Our framework makes it rather easy to do so, for the first
time.
\end{itemize}
In short, instead of seeing build systems as unrelated
points in space, we can re-envisage them as locations in a landscape,
leading to better understanding of what they do and how they compare,
and suggesting exploration of other (as yet unoccupied points) in the
landscape.

### 2018-03-13 paper/2-background.tex

Build systems automate the execution of simple repeatable tasks for individual
users, as well as for large organisations. There are software build systems,
such as \Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake} and
\Bazel~\cite{bazel}, as well as various incremental calculation engines, such
as \Excel~\cite{advanced_excel}. In this section we use these four examples to
introduce main domain-specific notions and requirements. Other notable examples
of build systems and their relation to these four are discussed
in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.

We start by giving an intuitive definition of a build system. After providing
more context and examples in the rest of this section, we
develop a formal definition in~\S\ref{sec-general-build}.

\definition[Build system]{Given a description of tasks, a \emph{build system}
executes them in an order that respects their dependencies to produce a
specified output from supplied inputs.}
\label{def-build}

@@

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and which is
not known statically. One might say that the value of \cmd{C1} is statically
known in this particular example, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    files <- lines <$> readFile "release.txt"
    need files
    system "tar" $ ["-cf", "result.tar"] ++ files
\end{minted}
\vspace{1mm}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting}~\cite{acar2007selfadjusting} \emph{computations} (see a
discussion in related work~\S\ref{sec-related}). Second, instead of abandoning
and deferring the execution of tasks whose newly discovered dependencies have
not yet been built (as \Excel does), \Shake \emph{pauses} their execution until
the dependencies are brought up-to-date. We refer to this build algorithm as
\emph{recursive}.

>>>>

Build systems automate the execution of simple repeatable tasks for individual
users, as well as for large organisations. In this section we explore the design
space of build systems, using four concrete examples:
\Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake},
\Bazel~\cite{bazel}, and \Excel~\cite{advanced_excel}\footnote{Excel appears
very different to the others but, seen through the lens of this paper, it is
very close indeed.}.
We have carefully chosen these four to illustrate the various axes on which build systems
differ; we discuss many other notable examples
of build systems, and their relationships,
in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.

@@

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and which is
not known statically\footnote{One might say that the value of \cmd{C1} is statically
known in this particular example, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    { need ["release.txt"]
    ; files <- lines <$> readFile "release.txt"
    ; need files
    ; system "tar" $ ["-cf", "result.tar"] ++ files }
\end{minted}
\vspace{1mm}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and \S\ref{sec-related}).
Second, instead of abandoning
and deferring the execution of tasks whose newly discovered dependencies have
not yet been built (as \Excel does), \Shake \emph{pauses} their execution until
the dependencies are brought up-to-date. We refer to this build algorithm as
\emph{recursive}.
